MADA Project - Ibis Blood Parasites from Florida

### 2021-10-07 products/manuscript/Manuscript.Rmd
Completed Project Part 2 10/7/21
Completed up to Part 2 of the MADA project. Data cleaning and exploratory analysis have been finished, along with all organization of the folders and files. Updated READMEs. Will return in later parts to data cleaning or exploratory analysis as needed.

---
title: "MADA Project - Ibis Blood Parasites from Florida"
subtitle: ""
author: Ryan Grunert
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
  bookdown::html_document2: 
    toc: false
bibliography: ../references.bib
csl: ../apa.csl
---
```{r}
#Loading required packages
library(dplyr) #for data processing
library(here) #to set paths
library(tidyverse) #all required data manipulation packages
library(xlsx) #to read in excel spreadsheets
```

@@

glimpse(rdIbisBlood15_17)
glimpse(rdIbisBlood10_14)

@@

glimpse(rdIbisField)

@@

#Loads the raw Ibis Urban Gradient supplemental data sheet
UrbanIbis <- read.xlsx(data_locationIbisUr, sheetName = "Samples by Date")

glimpse(UrbanIbis)

@@

glimpse(SalIbis)
```

@@

## Expected Analyses
I think that I will have to conduct multiple ANOVA's during the analysis process. Linear regression models may apply here too for some variables, but I'm not entirely sure. I would like to explore the possibility of using the location data for each sample to conduct a geographical analysis with a map as the output. I will be continuing to contact those involved with the study as well to see what analyses they want to be conducted and what questions they want answered.

# Methods and Results

## Data aquisition

## Data import and cleaning

@@

## Full analysis

# Discussion

## Summary and Interpretation

## Strengths and Limitations

## Conclusions

# References

# Summary/Abstract
_Write a summary of your project._

>>>>

---
title: "MADA Project - Ibis Blood Parasites from Florida"
subtitle: ""
author: Ryan Grunert
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
  bookdown::html_document2: 
    toc: false
bibliography: ../references.bib
csl: ../apa.csl
---
```{r}
#Loading required packages #for data processing
library(here) #to set paths
library(tidyverse) #all required data manipulation packages
library(xlsx) #to read in excel spreadsheets
library(janitor) #To assist in cleaning up dates and tidy data
library(lubridate) #To extract times from raw data
```

@@

str(rdIbisBlood15_17)
str(rdIbisBlood10_14)

@@

str(rdIbisField)
```

@@

#Loads the raw Ibis Urban Gradient supplemental data sheet
rdUrbanIbis <- read.xlsx(data_locationIbisUr, sheetName = "Samples by Date")

str(rdUrbanIbis)

@@

str(SalIbis)
```

@@

Is there a difference in HaeParasit between Ibis of different ages?
- Dependent variable (outcome): HaeParasit
- Independent variable (predictor): Age

 Is there a difference in HaeParasit between Ibis of different sex?
- Dependent variable (outcome): HaeParasit
- Independent variable (predictor): Sex

Is there a different in HaeParasit in Ibis sampled at different sites?
 - Dependent variable (outcome): HaeParasit
 - Independent variable (predictor): Site

 Is there a difference in HaeParasit between Ibis sampled in different habitat types?
 - Dependent variable (outcome): HaeParasit
 - Independent variable (predictor): HabType

 Is there a difference in HaeParasit between Ibis sampled at different times?
 - Dependent variable (outcome): HaeParasit
 - Independent variables (predictors): Date, Season

## Expected Analyses
I think that I will have to conduct multiple ANOVA's during the analysis process. Linear and logistic regression models apply here too for some variables.I would like to explore the possibility of using the location data for each sample to conduct a geographical analysis with a map as the output. I will be continuing to contact those involved with the study as well to see what analyses they want to be conducted and what questions they want answered.

## Data Cleaning
The data has already been imported, so now we move onto the cleaning stage. The following code will clean, process, and save the raw Ibis blood datasets from 2010-2014 and 2015-2017. This section is contained in the "ProcessingScriptIbisBloodData.R" file.
```{r}
##############################Cleaning the rdIbisBlood10_14 data###########################
IbisBlood10_14 <- rename(rdIbisBlood10_14, ID = X.) #renames the ID column

#Extracts just the columns of interest
IbisBlood10_14 <- select(IbisBlood10_14, ID, Site.ID, Date, PCR.sex, Age..adult.juvenile., 
                         Total.RBC, Haemoproteus.parasitemia.)

#Renames the columns of interest
IbisBlood10_14 <- rename(IbisBlood10_14, Site = Site.ID, Sex = PCR.sex, Age = Age..adult.juvenile.,
                         TotalRBC = Total.RBC, HaeParasit = Haemoproteus.parasitemia.)

#Change "adult" and "juvenile" in the Age column to "A" and "J"
IbisBlood10_14$Age[IbisBlood10_14$Age == "adult"] <- "A"
IbisBlood10_14$Age[IbisBlood10_14$Age == "Adult"] <- "A"
IbisBlood10_14$Age[IbisBlood10_14$Age == "juvenile"] <- "J"
IbisBlood10_14$Age[IbisBlood10_14$Age == "SA"] <- "A"
#Now the Ibises that had age recorded are referred to as "A" for adult or "J" for juvenile

#Changing the DIV/0 values in HaeParasit to NA and change the column to numerical values
IbisBlood10_14$HaeParasit <- as.numeric(IbisBlood10_14$HaeParasit)  # Divide-by-zero errors are now NA

#Reformatting the date values in the Date column
IbisBlood10_14$Date <- excel_numeric_to_date(IbisBlood10_14$Date)
#After this step, the values that just state the year are labelled with "1905", may need to go back or exclude from analysis

#Abbreviating the site names to make analysis easier
IbisBlood10_14$Site[IbisBlood10_14$Site == "Royal Palms"] <- "RP"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Everglades"] <- "E"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Dreher Park"] <- "DRP"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Indian Creek"] <- "ICP"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Indian Creek "] <- "ICP"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Indian Creek Park"] <- "ICP"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Juno Beach"] <- "JB"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Lion Country Safari"] <- "LCS"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Lake Worth"] <- "LW"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Prosperity Oaks"] <- "PO"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Promenade Plaza"] <- "PP"
IbisBlood10_14$Site[IbisBlood10_14$Site == "San Mateo"] <- "SM"
IbisBlood10_14$Site[IbisBlood10_14$Site == "Solid Waste Authority"] <- "SWA"
IbisBlood10_14$Site[IbisBlood10_14$Site == "West Palm Beach Zoo"] <- "WPBZ"

#Reformatting the mislabeled dates. These are set to Jan 1st of the year they were
#recorded on the data sheet, these entries only had the year stated, not month and day.
IbisBlood10_14$Date[IbisBlood10_14$Date == "1905-07-02"] <- "2010-01-01"
IbisBlood10_14$Date[IbisBlood10_14$Date == "1905-07-03"] <- "2011-01-01"
IbisBlood10_14$Date[IbisBlood10_14$Date == "1905-07-04"] <- "2012-01-01"

#The bottom of the dataset has a couple rows of just missing data for all variables,
#This code filters the dataset so that if any rows are present that has at least
#one variable be NOT NA, R keeps it
IbisBlood10_14 <- filter(IbisBlood10_14, if_any(everything(), ~ !is.na(.)))

#This completes the cleaning phase for IbisBlood10_14
#------------------------------------------------------------------------------------#

############################Cleaning the rdIbisBlood15_17 Data#########################

IbisBlood15_17<- rename(rdIbisBlood15_17, ID= X.) #renames the ID column

####Extracting only the columns relevant for analysis
IbisBlood15_17 <- select(IbisBlood15_17,ID,Site.Name,Date,PCR.sex,Age,Season,Habitat.type,
                         Total.RBC,X..Haemoproteus,Haemoproteus.parasitemia.,F1,F2,F3,
                         Avg.RBC.FOV,Ibis..,Total.mass..g.,Mass.bag..g.,Mass.bird..g.,Body.condition.score..1.5.,
                         Ectoparasite.score..1.5.,Culmen.length..mm.,Wing.chord.length..mm.,
                         Tarsus.length..mm.,Tarsus.width..mm.)

####Renaming the columns 
IbisBlood15_17 <- rename(IbisBlood15_17,Site = Site.Name,Sex = PCR.sex,HabType = Habitat.type,TotalRBC = Total.RBC,NumHae = X..Haemoproteus,
                         HaeParasit = Haemoproteus.parasitemia.,AvgRBC = Avg.RBC.FOV,
                         IbisNum = Ibis..,TotalMassG = Total.mass..g.,BagMassG = Mass.bag..g.,
                         BirdMassG = Mass.bird..g.,BodyCondScore = Body.condition.score..1.5.,
                         EctoParasitScore = Ectoparasite.score..1.5.,CulmenLmm = Culmen.length..mm.,
                         WingChordLmm = Wing.chord.length..mm.,TarsusLmm = Tarsus.length..mm.,
                         TarsusWmm = Tarsus.width..mm.)

####Site column
#Ibis that have sites matched with the 10_14 dataset will have the same labels
IbisBlood15_17$Site[IbisBlood15_17$Site == "Juno Beach"] <- "JB"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Indian Creek Park"] <- "ICP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Dubois park"] <- "DUP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Dreher Park"] <- "DRP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Lion Country Safari"] <- "LCS"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Dubois Park"] <- "DUP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Loxahatchee Wildlife Refuge"] <- "LOXWR"
IbisBlood15_17$Site[IbisBlood15_17$Site == "J.W. Corbett Wildlife Management Area"] <- "JWC"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Gaines Park"] <- "GP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Solid Waste Authority"] <- "SWA"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Fisheating creek"] <- "FC"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Kitching Creek"] <- "KC"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Loxahatchee NE"] <- "LOXNE"
IbisBlood15_17$Site[IbisBlood15_17$Site == "DuBois Park"] <- "DUP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "TetraTech"] <- "TT"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Green Cay"] <- "GC"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Dreher park"] <- "DRP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "CROW Wildlife Rehabilitation Center"] <- "CROW"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Lake Worth"] <- "LW"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Royal Palm"] <- "RP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "JW Corbett"] <- "JWC"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Busch Wildlife Sanctuary"] <- "BWS"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Loxahatchee Slough"] <- "LOXS"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Indian Creek"] <- "ICP"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Cat House (Bonnie's, Richard Lane)"] <- "CAT"
IbisBlood15_17$Site[IbisBlood15_17$Site == "Palm Beach Zoo"] <- "WPBZ"

####Date column
#Changing the date column to numeric values to allow date extraction
IbisBlood15_17$Date <- as.numeric(IbisBlood15_17$Date) 
##Reformatting the date values in the Date column
IbisBlood15_17$Date <- excel_numeric_to_date(IbisBlood15_17$Date)

####Sex column
IbisBlood15_17$Sex[IbisBlood15_17$Sex == "not bled"] <- NA
#Two values have missing data because the birds were not bled to determine sex
IbisBlood15_17$Sex[IbisBlood15_17$Sex == "M?"] <- NA
IbisBlood15_17$Sex[IbisBlood15_17$Sex == "F?"] <- NA
IbisBlood15_17$Sex[IbisBlood15_17$Sex == "F "] <- "F"
#Two values have question marks for unsure sex determination, removing those values

####Age column
#When an ibis reaches 4 years old, their plumage becomes all white. Any Ibis
#that was sampled with brown plumage is considered a juvenile. Since Haemoproteus is
#a chronic disease that mostly likely accumulates over time, long-lived birds should
#theoretically have more parasitemia. Could not be true though and may come back to
#look at the individual years in the juvenile birds
IbisBlood15_17$Age[IbisBlood15_17$Age == "Juvenile"] <- "J"
IbisBlood15_17$Age[IbisBlood15_17$Age == "Adult"] <- "A"
IbisBlood15_17$Age[IbisBlood15_17$Age == "2nd/3rd year"] <- "J"
IbisBlood15_17$Age[IbisBlood15_17$Age == "3rd year"] <- "J"
IbisBlood15_17$Age[IbisBlood15_17$Age == "1st year"] <- "J"
IbisBlood15_17$Age[IbisBlood15_17$Age == "adult"] <- "A"
IbisBlood15_17$Age[IbisBlood15_17$Age == "2nd year"] <- "J"
IbisBlood15_17$Age[IbisBlood15_17$Age == "NA"] <- NA
IbisBlood15_17$Age[IbisBlood15_17$Age == "2nd Year"] <- "J"
IbisBlood15_17$Age[IbisBlood15_17$Age == "Late third year"] <- "J"
IbisBlood15_17$Age[IbisBlood15_17$Age == "ADULT"] <- "A"
IbisBlood15_17$Age[IbisBlood15_17$Age == "Adult bc"] <- "A" #Not sure what bc stands for

#Season and HabType are fine the way they are

####TotalRBC
IbisBlood15_17$TotalRBC <- as.numeric(IbisBlood15_17$TotalRBC)
#DIV/0 errors are now NA in this column

#NumHae column is fine

####HaeParasit
IbisBlood15_17$HaeParasit <- as.numeric(IbisBlood15_17$HaeParasit)

####-999 values
IbisBlood15_17[IbisBlood15_17 == -999] <- NA

####Removing the last couple empty rows
IbisBlood15_17 <- IbisBlood15_17[-c(401, 402, 403, 404, 405, 406),]

#This completes the cleaning phase for IbisBlood15_17
#------------------------------------------------------------------------------------#

###############################Saving the Cleaned Datasets###########################

#Saving the Ibis Blood 10-14 dataset as an RDS in the processed data folder
save_data_location <- here::here("data", "processed_data", "processedIbisBlood10_14.rds")
saveRDS(IbisBlood10_14, file = save_data_location)

#Saving the Ibis Blood 15-17 dataset as an RDS in the processed data folder
save_data_location2 <- here::here("data", "processed_data", "processedIbisBlood15_17.rds")
saveRDS(IbisBlood15_17, file = save_data_location2)

This next section will clean, process, and save the raw Ibis Field dataset. This section is contained in the "ProcessingScriptIbisFieldData.R" file.

```{r}
######################### #Cleaning the rdIbisField Data###############################

IbisField <- rename(rdIbisField, ID = ID..NSFSITE...) #renames the ID column

#Selecting the columns of interest to this analysis from the datasheet
IbisField <- select(IbisField,ID,Site.Name,Latitude,Longitude,Date,PCR.Sex,Age,Season,
                    Habitat.type,Ibis.number,Mass.bird..g.,Body.condition.score..1.5.,
                    Ectoparasite.score..1.5.,Culmen.length..mm.,Wing.chord.length..mm.,
                    Tarsus.length..mm.,Tarsus.width..mm.,Habituation.score.of.flock..1.5.)

#Renaming the columns of interest, also matching the other cleaned datasets
IbisField <- rename(IbisField,Site = Site.Name,Lat = Latitude,Long = Longitude,Sex = PCR.Sex,
                    HabType = Habitat.type,IbisNum = Ibis.number,BirdMassG = Mass.bird..g.,
                    BodyCondScore = Body.condition.score..1.5.,EctoParasitScore = Ectoparasite.score..1.5.,
                    CulmenLmm = Culmen.length..mm.,WingChordLmm = Wing.chord.length..mm.,
                    TarsusLmm = Tarsus.length..mm.,TarsusWmm = Tarsus.width..mm.,HabFlockScore = Habituation.score.of.flock..1.5.)

#This dataset has "-999" values which mean they are intentionally left blank,
#Birds with an ID number of "-999" represent sites that didn't catch any birds,
#but had weather data collected. This code deletes those observations.
IbisField <- IbisField[!(IbisField$ID == "-999"),]

####Site
#These site location abbreviations are the same across datasets
IbisField$Site[IbisField$Site == "Busch Wildlife Sanctuary"] <- "BWS"
IbisField$Site[IbisField$Site == "Cat House (Bonnie's, Richard Lane)"] <- "CAT"
IbisField$Site[IbisField$Site == "Dubois Park"] <- "DUP"
IbisField$Site[IbisField$Site == "DuBois Park"] <- "DUP"
IbisField$Site[IbisField$Site == "Dreher Park"] <- "DRP"
IbisField$Site[IbisField$Site == "Fisheating Creek"] <- "FC"
IbisField$Site[IbisField$Site == "Green Cay"] <- "GC"
IbisField$Site[IbisField$Site == "Gaines Park"] <- "GP"
IbisField$Site[IbisField$Site == "Indian Creek"] <- "ICP"
IbisField$Site[IbisField$Site == "Juno Beach"] <- "JB"
IbisField$Site[IbisField$Site == "Juno Beach "] <- "JB"
IbisField$Site[IbisField$Site == "J.W. Corbett Wildlife Management Area"] <- "JWC"
IbisField$Site[IbisField$Site == "Kitching Creek"] <- "KC"
IbisField$Site[IbisField$Site == "Lion Country Safari"] <- "LCS"
IbisField$Site[IbisField$Site == "Lake Worth"] <- "LW"
IbisField$Site[IbisField$Site == "Loxahatchee Slough"] <- "LOXS"
IbisField$Site[IbisField$Site == "Loxahatchee Wildlife Refuge"] <- "LOXWR"
IbisField$Site[IbisField$Site == "Loxahatchee/LILA Cell B2"] <- "LOXB2"
IbisField$Site[IbisField$Site == "Loxahatchee NE "] <- "LOXNE"
IbisField$Site[IbisField$Site == "Palm Beach Zoo"] <- "WPBZ"
IbisField$Site[IbisField$Site == "Royal Palm"] <- "RP"
IbisField$Site[IbisField$Site == "CROW Wildlife Rehabilitation Center"] <- "CROW"
IbisField$Site[IbisField$Site == "Solid Waste Authority"] <- "SWA"
IbisField$Site[IbisField$Site == "TetraTech"] <- "TT"

####Sex Column
#This replaces the -999 and FAIL values with NA. The IbisBlood datasheets carry more sex data for the birds
IbisField$Sex[IbisField$Sex == "FAIL"] <- NA

####Age Column
IbisField$Age[IbisField$Age == "Adult"] <- "A"
IbisField$Age[IbisField$Age == "3rd year"] <- "J"
IbisField$Age[IbisField$Age == "2nd year"] <- "J"
IbisField$Age[IbisField$Age == "1st year"] <- "J"
IbisField$Age[IbisField$Age == "NA"] <- NA
IbisField$Age[IbisField$Age == "Adult "] <- "A"
IbisField$Age[IbisField$Age == "2nd year "] <- "J"
IbisField$Age[IbisField$Age == "Juvenile"] <- "J"
IbisField$Age[IbisField$Age == "2nd Year"] <- "J"

#Replacing all -999 values in the dataset with NA
IbisField[IbisField == -999] <- NA

#Replacing NA text strings with regular NA values in multiple columns
IbisField$BodyCondScore[IbisField$BodyCondScore == "NA"] <- NA
IbisField$HabFlockScore[IbisField$HabFlockScore == "NA"] <- NA
IbisField$CulmenLmm[IbisField$CulmenLmm == "NA"] <- NA
IbisField$WingChordLmm[IbisField$WingChordLmm == "NA"] <- NA
IbisField$TarsusLmm[IbisField$TarsusLmm == "NA"] <- NA
IbisField$TarsusWmm[IbisField$TarsusWmm == "NA"] <- NA

#--------------------------------------------------------------------------------------#

###############################Saving the Cleaned Dataset###############################

save_data_location3 <- here::here("data", "processed_data", "processedIbisFielddata.rds")
saveRDS(IbisField, file = save_data_location3)
```

This cleaning code section will clean, process, and save the Ibis urbanization dataset. This section is contained in the "ProcessingScriptIbisUrban.R" file.

```{r}

##############################Cleaning the rdUrbanIbis data###########################
#Selects only the columns of interest in the dataset
IbisUrban <- select(rdUrbanIbis, Name:Serotype)

#Renames the columns of interest
IbisUrban <- rename(IbisUrban,ID = Name,Date = Collection.Date,HgPPM = mg.kg.Hg..PPM.,
                    Site = Site.Name,UrbanPercent = Site...Urbanized)

#Change the site names to their abbreviations
IbisUrban$Site[IbisUrban$Site == "Juno Beach"] <- "JB"
IbisUrban$Site[IbisUrban$Site == "Indian Creek"] <- "ICP"
IbisUrban$Site[IbisUrban$Site == "Dubois Park"] <- "DUP"
IbisUrban$Site[IbisUrban$Site == "Dreher Park"] <- "DRP"
IbisUrban$Site[IbisUrban$Site == "Lion Country Safari"] <- "LCS"
IbisUrban$Site[IbisUrban$Site == "Loxahatchee Wildlife Refuge"] <- "LOXWR"
IbisUrban$Site[IbisUrban$Site == "Solid Waste Authority "] <- "SWA"
IbisUrban$Site[IbisUrban$Site == "Gaines Park"] <- "GP"
IbisUrban$Site[IbisUrban$Site == "Kitching Creek"] <- "KC"
IbisUrban$Site[IbisUrban$Site == "Kitching Creek "] <- "KC"
IbisUrban$Site[IbisUrban$Site == "Loxahatchee NE"] <- "LOXNE"
IbisUrban$Site[IbisUrban$Site == "TetraTech"] <- "TT"
IbisUrban$Site[IbisUrban$Site == "Loxahatchee NE "] <- "LOXNE"
IbisUrban$Site[IbisUrban$Site == "Green Cay"] <- "GC"
IbisUrban$Site[IbisUrban$Site == "J.W. Corbett Wildlife Management Area"] <- "JWC"
IbisUrban$Site[IbisUrban$Site == "Loxahatchee Slough"] <- "LOXS"
IbisUrban$Site[IbisUrban$Site == "Lake Worth"] <- "LW"

#--------------------------------------------------------------------------------------#

################################Saving the UrbanIbis Data#############################
#Saving the Ibis Blood 15-17 dataset as an RDS in the processed data folder
save_data_location3 <- here::here("data", "processed_data", "processedIbisUrban.rds")
saveRDS(IbisUrban, file = save_data_location3)

#--------------------------------------------------------------------------------------#
```

This last cleaning section combines the data sheets into one to use for analysis. The urbanization gradient has not been created as of yet, but a meeting this month with those involved with the project will help shed some light on the exact specifics for analysis. This code is located in the "ExploratoryAnalysisScript.R" file.

```{r}
###############################Completing Combination Data Frames#########################
#Outer joining IbisBlood10_14 and IbisBlood15_17, keeping all values of the 10_14 dataset
IbisBlood10_17 <- merge(IbisBlood10_14, IbisBlood15_17, by = c("ID", "Site", "Date", "Sex", "Age",
                                                              "TotalRBC", "HaeParasit"), all=TRUE)
#There is no overlap in the ID numbers between the two, so no recapture is assumed.

#This is a left join, this data frame below only contains birds that were in the IbisBlood10_17
#dataset (all with parasitemia data), with additional information added by the IbisField data. Ibis
#from the Field data that weren't tested for parasitemia or weren't included on Blood dataset
#are not included in this dataframe

#IbisBlood10_17(with Field)_(Parasitemia data)
IbisC <- left_join(IbisBlood10_17, IbisField, by = c("ID", "Site", "Date", "Sex", "Age", "Season", "HabType",
                                                         "IbisNum", "BirdMassG", "BodyCondScore", "EctoParasitScore",
                                                         "CulmenLmm", "WingChordLmm", "TarsusLmm", "TarsusWmm"))

#This section classifies the habitat type for the sites without a habitat type.
IbisC <- IbisC %>%
        mutate(HabType = ifelse(Site == "ICP", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "JB", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "RP", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "LCS", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "LW", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "SWA", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "E", "Natural", HabType)) %>%
        mutate(HabType = ifelse(Site == "WPBZ", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "PO", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "PP", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "DRP", "Urban", HabType)) %>%
        mutate(HabType = ifelse(Site == "SM", "Urban", HabType))

##This section is adding in the seasons for observations that have collection dates but no season.

#Creating intervals for adding seasons based on the date column
int1 <- interval(ymd("2012-12-01"), ymd("2012-12-31"))
int2 <- interval(ymd("2013-06-01"), ymd("2013-08-31"))
int3 <- interval(ymd("2013-12-01"), ymd("2013-12-31"))
int4 <- interval(ymd("2014-02-01"), ymd("2014-03-31"))

#Adding seasons to the ibis without them stated based on the date the sameple was collected
IbisC <- IbisC %>%
         mutate(Season = ifelse(Date %within% int1, "Fall 2012", Season)) %>%
         mutate(Season = ifelse(Date %within% int2, "Summer 2013", Season)) %>%
         mutate(Season = ifelse(Date %within% int3, "Fall 2013", Season)) %>%
         mutate(Season = ifelse(Date %within% int4, "Spring 2014", Season))

#Adding a column for HaeParasit Presence/Absence data
IbisC <- IbisC %>%
    mutate(HaeParasitPA = as.integer(HaeParasit > 0 & !is.na(HaeParasit)))

#Adding a column for transformed HaeParasit data on a logarithm base 10 scale,
#then removing the -Inf values due to there being no parasitemia in those birds. This column
#then only contains logarithm transformed parasitemia data for birds with a positive value
#of parasitemia.
IbisC <- IbisC %>%
    mutate(HaeParasitLog10 = log10(HaeParasit)) %>%
    mutate(HaeParasitLog10 = ifelse(HaeParasitLog10 == "-Inf", NA, HaeParasitLog10))

#Changing the BodyCondScore to an ordinal scale
IbisC <- IbisC %>%
      mutate(BodyCondScore = recode_factor(BodyCondScore, '1' = 'Emaciated',
                                           '2' = 'Underweight', '3' = 'Ideal',
                                           '4' = 'Overweight', '5' = 'Obese',
                                           .ordered = TRUE))

#Changing the class of some variables for analysis. From character to numeric.
C2N <- c(16,17,18,21,22,23,24)

IbisC[ , C2N] <- apply(IbisC[ , C2N], 2,
                       function(x) as.numeric(as.character(x)))

This concludes the data cleaning section, we are now ready for the exploratory analysis!

@@

### Tables

To start the exploratory analysis, we're going to create some quick tables to get an idea of the metrics for some of the columns of interest.

```{r}
IbisC %>%
  count(Sex, wt = HaeParasitPA)
```

Looking at presence/absence data for haemoproteus parasitemia, there were about double the number of female ibises with active infection compared to the males. 78 of the ibises sampled were negative for haemoproteus infection.

```{r}
IbisC %>%
  count(Age, wt = HaeParasitPA)
```

Here we can see that there was about 2.7 times more adults sampled with active infection than juvenile ibises. 59 of the sampled ibises with active infection either couldn't be aged or age was not recorded.

```{r}
IbisC %>%
  count(Age, Sex, wt = HaeParasitPA)
```

Combining age and sex, we can see that adult females had the highest number of active infections. Categories with missing numbers are prevalent.

```{r}
IbisC %>%
  count(Site, wt = HaeParasitPA)
```
Looking at site, SWA, ICP, and DRP were the sites with the highest number of sampled ibises with active infection.

```{r}
IbisC %>%
  count(Season, wt = HaeParasitPA)
```
Looking at season sampled, Spring 2016 had the highest number of sampled ibises with active infection, followed by Summer 2013, Spring 2017, and Spring 2014.

```{r}
IbisC %>%
  count(BodyCondScore, Age, wt = HaeParasitPA)
```
Looking at body condition score, the majority of the birds sampled with active infection were an ideal weight or underweight. This lets us know that an active infection didn't affect body condition too harshly on these birds.

### Figures

Now we can move onto creating some figures for the data to see any patterns visually.

Figure 1: Haemoproteus parasitemia from sampled infected Ibises from 2010-2017 by habitat type
```{r}
ExpFigure1 <-ggplot(IbisC) +
  geom_point(aes(x = Date, y = HaeParasitLog10, color = HabType)) +
  ggtitle("Haemoproteus parasitemia from sampled infected Ibises from 2010-2017 by habitat type") +
  xlab("Date") +
  ylab("Haemoproteus Parasitemia (Log base 10)") +
  scale_color_discrete("Habitat Type", na.translate = F) +
  theme(plot.title = element_text(hjust = 0.5, size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        legend.title = element_text(size = 10))

Figure_file1 <- here("results", "ExploratoryAnalysis", "ExpFigure1.png")
ggsave(filename = Figure_file1,plot = ExpFigure1)

ExpFigure1
```

This scatterplot graphs the sample dates chronologically against the parasitemia measured from the sampled ibis on a transformed log base 10 scale. Habitat type is depicted by color of the dots. Overall, there is a general narrowing of the data over time with a slight negative correlation. There isn't a clear difference between natural and urban habitat types when it comes to amount of parasitemia.

Figure 2: Haemoproteus parasitemia from adult and juvenile sampled Ibises
```{r}
ExpFigure2 <- IbisC %>%
  drop_na(Age) %>%
  ggplot() +
  geom_violin(aes(x = Age, y = HaeParasit)) +
  ggtitle("Haemoproteus parasitemia from adult and juvenile sampled Ibises") +
  xlab("Age") +
  ylab("Haemoproteus Parasitemia") +
  scale_y_continuous(limits = c(0, 0.1)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        legend.title = element_text(size = 10))

Figure_file2 <- here("results", "ExploratoryAnalysis", "ExpFigure2.png")
ggsave(filename = Figure_file2, plot = ExpFigure2)

ExpFigure2
```

This violin plot graphs the haemoproteus parasitemia distribution for the adult and juvenile Ibises sampled. The majority of the parasitemia values are on the lower end for both adults and juveniles. The distribution for adults reaches higher than juveniles. The overall structure for the adult distribution is higher and skinnier, while juveniles have more of their parasitemias lower in value.

Figure 3: Haemoproteus parasitemia from sampled Ibises by sex
```{r}
ExpFigure3 <- IbisC %>%
  drop_na(Sex) %>%
  ggplot() +
  geom_boxplot(aes(x = Sex, y = HaeParasit)) +
  scale_y_continuous(limits = c(0, 0.5)) +
  ggtitle("Haemoproteus parasitemia from sampled Ibises by sex") +
  xlab("Sex") +
  ylab("Haemoproteus Parasitemia") +
  theme(plot.title = element_text(hjust = 0.5, size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        legend.title = element_text(size = 10))

Figure_file3 <- here("results", "ExploratoryAnalysis", "ExpFigure3.png")
ggsave(filename = Figure_file3, plot = ExpFigure3)

ExpFigure3
```

The haemoproteus parasitemia distributions between sexes is extremely similar. The median value for males is slightly higher than females. The majority of the measured parasitemias are between 0.0 and 0.1.

Figure 4: Haemoproteus parasitemia by total female and male Ibis mass
```{r}
ExpFigure4 <- IbisC %>%
  drop_na(BirdMassG, Sex) %>%
  ggplot() +
  geom_point(aes(x = BirdMassG, y = HaeParasitLog10, color = Sex)) +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_continuous(limits = c(130, 1240)) +
  ggtitle("Haemoproteus parasitemia by total female and male Ibis mass") +
  xlab("Total mass of Ibis (g)") +
  ylab("Haemoproteus Parasitemia (Log base 10)") +
  theme(plot.title = element_text(hjust = 0.5, size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        legend.title = element_text(size = 10))

Figure_file4 <- here("results", "ExploratoryAnalysis", "ExpFigure4.png")
ggsave(filename = Figure_file4, plot = ExpFigure4)

ExpFigure4
```

Looking at the plot, male birds seem to weigh more than females. However, when comparing the parasitemia between females and males, there isn't a clear difference between the two.

Figure 5: Haemoproteus parasitemia by female and male Ibis mass and body condition score
```{r}
ExpFigure5 <- IbisC %>%
  drop_na(BirdMassG, BodyCondScore, Sex) %>%
  ggplot() +
  geom_point(aes(x = BirdMassG, y = HaeParasitLog10, color = BodyCondScore)) +
  facet_wrap(~ Sex) +
  ggtitle("Haemoproteus parasitemia by female and male Ibis mass and body condition score") +
  xlab("Total mass of Ibis (g)") +
  ylab("Haemoproteus Parasitemia (Log base 10)") +
  theme(axis.text.x = element_text(angle = 90),
        plot.title = element_text(hjust = 0.5, size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        legend.title = element_text(size = 10)) +
  scale_x_continuous(limits = c(130, 1240)) +
  scale_color_discrete("Body Condition", na.translate = F)

Figure_file5 <- here("results", "ExploratoryAnalysis", "ExpFigure5.png")
ggsave(filename = Figure_file5, plot = ExpFigure5)

ExpFigure5
```

Looking at this plot, there are more underweight female ibises with higher parasitemias compared to males. This could be due to the higher number of sampled females though.
